---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r load_and_preprocess}
## Install and load libraries
.libPaths("c:/temp/rpackages")
suppressWarnings({
  # install.packages(c("data.table", "knitr", "ggplot2"))
  library("data.table")
  library("knitr")
  library(ggplot2)
})

opts_chunk$set(echo=TRUE)

# Loading and preprocessing the data
unzip("activity.zip")
remove(activity)
activity <- read.csv("activity.csv", na.strings = "NA", header = T)
activity[,"date"] <- as.Date(activity$date)
activity[,"DateTime"] <-
  as.POSIXct(paste(activity$date, sprintf("%04d",activity$interval)),
             "%Y-%m-%d %H%M", tz="GMT")
# activity <- as.data.table(activity)
# setkey(activity, c("date", "interval"))
```

## What is mean total number of steps taken per day?
```{r steps_per_day}
steps_per_day = aggregate(. ~ date, data = activity, FUN=sum)[,c("date", "steps")]
ggplot(steps_per_day, aes(x=date, weights=steps)) +
  geom_histogram(aes(fill = ..count..),binwidth = 1) +
  scale_fill_gradient(high="green", low="red")

average_steps_per_day = list(mean = mean(steps_per_day$steps),
                             median = median(steps_per_day$steps))
average_steps_per_day
```



## What is the average daily activity pattern?
```{r daily_activity}
steps_per_interval = aggregate(. ~ interval, data = activity, FUN=mean)[,c("interval", "steps")]
plot(y=steps_per_interval$steps, x=steps_per_interval$interval, type='l',
     ylab = "Steps per 5min interval", xlab = "Minute of day",
     main="Average Step Count by Time of Day", col = "red")

best_interval <- steps_per_interval[which.max(steps_per_interval$steps),]
```
The best time of day interval is `r best_interval$interval` with `r as.integer(best_interval$steps)` steps on average.

## Imputing missing values

Some of the data is missing (not available).  This section replaces missing data with the average for that interval.

```{r imputed_steps_per_day}

for(i in 1:nrow(activity)){
  interval <- activity[i, "interval"]
  avg_steps <- steps_per_interval[which(steps_per_interval$interval == interval), "steps"][1]
  activity[i, "imputed_steps"] <-
    if(is.na(activity[i, "steps"]))
      avg_steps
    else
      activity[i, "steps"]
}

imputed_steps_per_day = aggregate(. ~ date, data = activity[,c("date", "imputed_steps")], FUN=sum)[,c("date", "imputed_steps")]
ggplot(imputed_steps_per_day, aes(x=date, weights=imputed_steps)) +
  geom_histogram(aes(fill = ..count..),binwidth = 1) +
  scale_fill_gradient(high="green", low="red")

imputed_average_steps_per_day = list(mean = mean(imputed_steps_per_day$imputed_steps),
                                     median = median(imputed_steps_per_day$imputed_steps))
imputed_average_steps_per_day
```

The mean values between the observed data and imputed data are the same.  The median value of the imputed data is the same as the mean.

## Are there differences in activity patterns between weekdays and weekends?

*Note:* This section uses imputed data.

```{r weekday_vs_weekend_activity}

imputed_steps_per_interval_by_day_of_week <- function(days){
  aggregate(. ~ interval,
            data = activity[weekdays(activity$DateTime) %in% days,],
            FUN=mean)[,c("interval", "imputed_steps")]
}

# Weekend
weekend_steps_per_interval <- imputed_steps_per_interval_by_day_of_week(
  c('Saturday','Sunday'))
weekend_best_interval <- weekend_steps_per_interval[which.max(weekend_steps_per_interval$imputed_steps),]

# Weekday
weekday_steps_per_interval <- imputed_steps_per_interval_by_day_of_week(
  c('Monday','Tuesday',"Wednesday","Thursday","Friday"))

weekday_best_interval <- weekday_steps_per_interval[which.max(weekday_steps_per_interval$imputed_steps),]

# Plot 
par(mfrow = c(2, 1), mar = c(4,4,2,1), oma = c(0,0,2,0))

plot(y=weekend_steps_per_interval$imputed_steps, x=weekend_steps_per_interval$interval, type='l',
     ylab = "Steps per 5min", xlab = "",
     main="Weekend", col = "red")

plot(y=weekday_steps_per_interval$imputed_steps, x=weekday_steps_per_interval$interval, type='l',
     ylab = "Steps per 5min", xlab = "Minute of day",
     main="Weekday", col = "blue")
```

The best weekend time of day interval is `r weekend_best_interval$interval` with `r as.integer(weekend_best_interval$imputed_steps)` steps on average.

The best weekday time of day interval is `r weekday_best_interval$interval` with `r as.integer(weekday_best_interval$imputed_steps)` steps on average.

Here the charts are overlaid:

```{r weekday_overlaid_weekend_activity}
plot(y=weekday_steps_per_interval$imputed_steps, x=weekday_steps_per_interval$interval,
     ylab = "Steps per 5min interval", xlab = "Minute of day",
     main="Weekday vs. Weekend",
     type='l', col = "blue")

lines(y=weekend_steps_per_interval$imputed_steps, x=weekend_steps_per_interval$interval,
      type='l', col = "red")

legend("topleft",
       legend=c("Weekday","Weekend"),
       col = c("Blue","Red"),
       bty = "n",
       pch = "_")

```

## Summary

From these images, we can infer that the individual rises 2-3 hours earlier on the weekdays than the weekend and often stays active later into the evening on the weekends.